### 2022-04-07 manuscript.tex
format [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/da43cb74d176c31bc2ce988d80e925e8ab7b75fa.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/da43cb74d176c31bc2ce988d80e925e8ab7b75fa/checks
https://github.com/scherzerthesis/thesis/actions/runs/2111157431

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/ca51cc941ab662e7bdc8ce017ea179a06238ba3f/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/ca51cc941ab662e7bdc8ce017ea179a06238ba3f}{scherzerthesis/thesis@ca51cc9}
on April 7, 2022.

@@

Genetic sequencing of human lung adenocarcinomas has also revealed the widespread inactivation of p53 as a common alteration in late-stage lung adenocarcinoma. Normal p53 function restricts tumor plasticity, and thus losing p53 function enables tumor cells to adopt more malignant cell identities. However, not all p53 mutations are created equal. In chapter 2, we characterized the effect of three p53 mutant alleles (P53-null, P53-R172H, and p53-R245W) on BRAF\textsuperscript{V600E} lung tumorigenesis. It was found that p53-R172H and p53-R245W induce more widespread tumor dedifferentiation, thus leading to worst tumor outcomes compared to p53-loss. Cell lines derived from the tumors allowed for the dissection of signaling pathway dynamics, which led us to observe higher MAPK pathway activation in p53-R172H and p53-R245W cell lines. We also found that p53-R172H and p53-R245W cell lines were more refractory to MAPK pathway inhibition with small molecule inhibitors. These altered phenotypes seen in p53-R172H and p53-R245W are likely a reflection of their altered transcriptomic profiles.

@@

The 5-year survival rate for patients with lung cancer was 25\% in 2020 but varied depending on the stage of disease at the time of diagnosis (Siegel et al., 2021). Like other solid cancers, lung cancers are most deadly after primary tumors have metastasized to other organ systems like the liver, kidney, bone, or nervous system (Matsunaga and Shida, 1971). Lung cancer patients with no local or distant metastasis typically have a 5-year survival percentage of approximately 40\% compared to approximately 5\% with distant metastasis{[}Woodard et al.; 27740970{]}. Ultimately, the cause of death of many lung cancer patients includes primary lung tumor burden, infection, pulmonary embolism or hemorrhage, or complications of metastatic dissemination(Nichols et al., 2012). Although patient survival has improved due to improved targeted and immune therapies, the still poor prognosis reflects a need to understand better the molecular mechanisms underlying lung adenocarcinoma initiation, progression, maintenance, and response to targeted- or immune therapy.

@@

Early on in P53 studies, there were reports of specific P53 mutations having a "gain-of-function" (GOF) effect. For example, Li-Fraumeni patients with specific missense mutations in P53 would develop tumors earlier than Li-Fraumeni patients with loss-of-function (LOF) mutations (Bougeard et al., 2008). Furthermore, experimental cell biology studies would express mutant-p53 in P53-null cells and demonstrate enhanced tumorigenic potential (Levine, 2020). Further evidence for GOF mutant-P53 in mice indicated that missense mutant P53 induced different cancer types and enhanced metastasis than LOF P53 (Olive et al., 2004). Furthermore, many groups have shown that mutant-p53 can alter signal transduction affecting chemoresistance and metabolism.

@@

We also failed to detect robust differences between genotypes known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely due to a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow complete transcriptomic changes to take effect imparted by tumor suppressor or oncogene mutation. Therefore, a time-course analysis that spans early time points (4-6 weeks) and late time points (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated no tumor growth advantage at early time points as late as six weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preparation. Specifically, there are gene-dropout effects where the droplet-based technique captures only the most highly expressed genes. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows for profiling more cells than similar scRNA-seq techniques. Genes with lower expression can be as important as highly expressed genes for a cell's identity. Transcription factors that determine cell states have been shown to have modest expression{[}Wang et al. (2021); (\textbf{pmid?});30205475{]}.
Therefore, we hypothesize that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow more gene capture. Thus, Smart-seq2 will enable us to determine transcriptional heterogeneity and genotype-specific gene expression changes more precisely.

@@

@@

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/da43cb74d176c31bc2ce988d80e925e8ab7b75fa/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/da43cb74d176c31bc2ce988d80e925e8ab7b75fa}{scherzerthesis/thesis@da43cb7}
on April 7, 2022.

@@

Genetic sequencing of human lung adenocarcinomas has also revealed the widespread inactivation of p53 as a common alteration in late-stage lung adenocarcinoma. Normal p53 function restricts tumor plasticity, and thus losing p53 function enables tumor cells to adopt more malignant cell identities. However, not all p53 mutations are created equal. In chapter 2, we characterized the effect of three p53 mutant alleles (P53\textsuperscript{fl/fl}, P53\textsuperscript{R172H}, and p53\textsuperscript{R245W}) on BRAF\textsuperscript{V600E} lung tumorigenesis. It was found that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} induce more widespread tumor dedifferentiation, thus leading to worst tumor outcomes compared to p53-loss. Cell lines derived from the tumors allowed for the dissection of signaling pathway dynamics, which led us to observe higher MAPK pathway activation in p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cell lines. We also found that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cell lines were more refractory to MAPK pathway inhibition with small molecule inhibitors. These altered phenotypes seen in p53\textsuperscript{R172H} and p53\textsuperscript{R245W} are likely a reflection of their altered transcriptomic profiles.

@@

The 5-year survival rate for patients with lung cancer was 25\% in 2020 but varied depending on the stage of disease at the time of diagnosis (Siegel et al., 2021). Like other solid cancers, lung cancers are most deadly after primary tumors have metastasized to other organ systems like the liver, kidney, bone, or nervous system (Matsunaga and Shida, 1971). Lung cancer patients with no local or distant metastasis typically have a 5-year survival percentage of approximately 40\% compared to approximately 5\% with distant metastasis(Duggan et al., 2016; Woodard et al.). Ultimately, the cause of death of many lung cancer patients includes primary lung tumor burden, infection, pulmonary embolism or hemorrhage, or complications of metastatic dissemination(Nichols et al., 2012). Although patient survival has improved due to improved targeted and immune therapies, the still poor prognosis reflects a need to understand better the molecular mechanisms underlying lung adenocarcinoma initiation, progression, maintenance, and response to targeted- or immune therapy.

@@

Early on in P53 studies, there were reports of specific P53 mutations having a "gain-of-function" (GOF) effect. For example, Li-Fraumeni patients with specific missense mutations in P53 would develop tumors earlier than Li-Fraumeni patients with loss-of-function (LOF) mutations (Bougeard et al., 2008). Furthermore, experimental cell biology studies would express mutant-p53 in P53\textsuperscript{null} cells and demonstrate enhanced tumorigenic potential (Levine, 2020). Further evidence for GOF mutant-P53 in mice indicated that missense mutant P53 induced different cancer types and enhanced metastasis than LOF P53 (Olive et al., 2004). Furthermore, many groups have shown that mutant-p53 can alter signal transduction affecting chemoresistance and metabolism.

@@

We also failed to detect robust differences between genotypes known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely due to a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow complete transcriptomic changes to take effect imparted by tumor suppressor or oncogene mutation. Therefore, a time-course analysis that spans early time points (4-6 weeks) and late time points (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated no tumor growth advantage at early time points as late as six weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preparation. Specifically, there are gene-dropout effects where the droplet-based technique captures only the most highly expressed genes. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows for profiling more cells than similar scRNA-seq techniques. Genes with lower expression can be as important as highly expressed genes for a cell's identity. Transcription factors that determine cell states have been shown to have modest expression(Martin and Sung, 2018; Wang et al., 2021).
Therefore, we hypothesize that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow more gene capture. Thus, Smart-seq2 will enable us to determine transcriptional heterogeneity and genotype-specific gene expression changes more precisely.

@@

\leavevmode\vadjust pre{\hypertarget{ref-iZbc8sIr}{}}%
Duggan, M.A., Anderson, W.F., Altekruse, S., Penberthy, L., and Sherman, M.E. (2016). The Surveillance, Epidemiology, and End Results (SEER) Program and Pathology: Toward Strengthening the Critical Relationship. Am J Surg Pathol \emph{40}, e94--e102. \url{https://doi.org/10.1097/pas.0000000000000749}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-YdeeJclQ}{}}%
Martin, E.W., and Sung, M.-H. (2018). Challenges of Decoding Transcription Factor Dynamics in Terms of Gene Regulation. Cells \emph{7}. \url{https://doi.org/10.3390/cells7090132}.
